---
title: 'Cyberville Dataset Description'
output:
  html_document:
    theme: united
css: laurens_styles.css
---

```{r, child="_Global-Options.Rmd"}
```
***
***
### families.csv

```{r}
tempData = read.csv("datasets/families.csv")

summary(tempData)
```

The data set families contains information about 43,886 families living in the city of Cyberville. The city has four regions. In these exercises, we will learn about the families of Cyberville by using sampling. For each family, the following information is recorded:

1.  Family type\
    **1:** Husband-wife family **2:** Male-head family **3:** Female-head family\
2.  Number of persons in family\
3.  Number of children in family\
4.  Family income\
5.  Region\
    **1:** North (10,149 families) **2:** East (10,390) **3:** South (13,457) **4:** West (9,890)\
6.  Education\
    31: Less than 1st grade\
    32: 1st, 2nd, 3rd, or 4th grade\
    33: 5th or 6th grade\
    34: 7th or 8th grade\
    35: 9th grade\
    36: 10th grade\
    37: 11th grade\
    38: 12th grade, no diploma\
    39: High school graduate, high school diploma, or equivalent\
    40: Some college but no degree\
    41: Associate degree in college (occupation/vocation program)\
    42: Associate degree in college (academic program)\
    43: Bachelor's degree (e.g., B.S., B.A., A.B.)\
    44: Master's degree (e.g., M.S., M.A., M.B.A.)\
    45: Professional school degree (e.g., M.D., D.D.S., D.V.M., LL.B., J.D.)\
    46: Doctoral degree (e.g. Ph.D., Ed.D.)  
```{r}

```
    
    
