---
title: 'Tutorial 1: Working with data'
author: "Professor Lauren Cipriano"
date: "2024-05-05"
output:
  html_document:
    theme: united
    toc: true
    toc_float: true
css: laurens_styles.css
---

```{r, include=FALSE}
# Setting default options for each chunk of code
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

------------------------------------------------------------------------

## Part 1. Starting with Data

### Step 1. Import the datafile

------------------------------------------------------------------------

You can import data into R using the import wizard, but often it is useful to change the working directory and import the data file using a line of code.

One advantage of using a code line to import your data is that it is easily reproducible.

```{r, eval=FALSE, echo=FALSE}
setwd("/home/ian/Documents/GitHub/Cloud-Folder/Lauren-Work/Lauren Rmarkdown")
```

```{r, eval=FALSE}
# Set the working directory
setwd("C:/Users/Lauren/Dropbox/Ivey/2024 MSc Statistics/Labs/")
```

```{r}
# Import the data - as a csv file
data = read.csv("datasets/families.csv")
```

### Step 2. View your data

------------------------------------------------------------------------

```{r, eval=FALSE}
## Option 1. Open the dataset in another window.  It will look like a spreadsheet
View(data)
```

```{r, eval=FALSE}
## Option 2. Call the dataset to print it in the coding window
data
```

```{r}
## Option 3. Ask to see a section of the data 
# just the first 6 rows 
head(data)

# just the last 6 rows 
tail(data)
```

### Step 3. Quick report on data

------------------------------------------------------------------------

It is important to know how R sees each of your variables. Asking for a summary of the data will provide summary statistics on the data.frame.

```{r}
summary(data)
```

### Step 4. Assign appropriate variable types

------------------------------------------------------------------------

**Character:** free text, a.k.a. string\
**Factor:** Categorical values; actual values may be text (North/South/East/West) or numbers (1/2/3/4)\
**Logical:** Binary (TRUE/FALSE)\
**Integer:** Whole numbers (1L/2L/3L/4L, the letter 'L' declares these as Integers)\
**Numeric:** Decimal Numbers (1.5/2.5/3.5/4.5)

#### Step 4a. Investigating types

```{r}
# Check the variable class of TYPE
class(data$TYPE)

# See how many unique values TYPE has
unique(data$TYPE)

# Tell R that the TYPE column can only have 3 possible values (turn it into a factor)
data$TYPE = factor(data$TYPE, levels = c(1,2,3))

# Now check the class of TYPE
class(data$TYPE)

# If we try to assign an invalid value to row 1, column "TYPE"
data[1, "TYPE"] = 4

# Warning: this assigns the value as NA, but lets you know you've made a mistake
data[1, "TYPE"]

# Change it back to integer
data$TYPE = as.integer(data$TYPE)

class(data$TYPE)

# Fixing NA from above
data = read.csv("datasets/families.csv")
```

#### Step 4b. Assigning New Types

```{r}
# Turning TYPE into a factor with easy-to-understand labels
data$TYPE = factor(data$TYPE, levels = c(1,2,3), labels = c("Husband-wife family",
                                                            "Male-head family",
                                                            "Female-head family"))

# Number of persons and children, and income are already integers
class(data$PERSONS); class(data$CHILDREN); class(data$INCOME)

# Turn REGION into a factor
data$REGION = factor(data$REGION, levels = c(1,2,3,4), label = c("NORTH",
                                                                 "EAST",
                                                                 "SOUTH",
                                                                 "WEST"))

# New factor for education

# Turn Education Level into a factor, might have not been worth it...
#data$EDUCATION = factor(data$EDUCATION, levels = c(31:46), label = c("<1st grade", "1-4th grade", "5,6th grade", "7,8th grade", "9th grade", "10th grade", "11th grade", "12th grade, no diploma", "High school graduate or equivalent", "Some college but no degree", "Associate degree in college (occupation/vocation program)", "Associate degree in college (academic program)", "Bachelor's Degree", "Master's Degree", "Professional School Degree", "Doctoral Degree"))

# Now we can view the summary of our data again
summary(data)
```

***

## Part 2. Analyzing Data

### Step 1. Some Summary Statistics
#### a) Proportion of female-headed families
```{r}
# Number of Families in the dataset
n_fam = nrow(data)

# The which function returns a vector that tells us which indices of a matrix are true given a condition, in this case female-headed
indices = which(data$TYPE=="Female-head family")

n_female_fam = nrow(data[indices,])

# Calculating the proportion
prop_female_fam = n_female_fam/n_fam
prop_female_fam
```
#### b) Average number of children per family
```{r}
# Total Number of children in all families
n_total_children = sum(data$CHILDREN)

# Recall we have n_total_fam from a)
avg_n_children = n_total_children/n_fam
avg_n_children

# or...
mean(data$CHILDREN)
```
#### c) Standard deviation of the number of children per family
```{r}
# R has a function for that!
sd(data$CHILDREN)
```

#### d) Proportion of heads of household who did not receive a high school diploma  
First we must look at our EDUCATION variable to see that any value coded below 39 means that no diploma was received
```{r}
# Using the which function again (note this will overwrite our previous indices variable since we chose the same name)
indices = which(data$EDUCATION<39)

# Counting the rows
n_no_diploma = nrow(data[indices,])

# Finding the proportion
prop_no_diploma = n_no_diploma/n_fam
prop_no_diploma
```

#### e) Proportion of families whos head received a bachelor's or master's degree and have one child
```{r}
# This time, we will combine multiple arguments in the which function using logical operators (| = OR, & = AND, ! = NOT)
indices = which((data$EDUCATION==43 | data$EDUCATION==44) & data$CHILDREN==1)

n_fam_of_interest = nrow(data[indices,])

prop_fam_of_interest = n_fam_of_interest/n_fam
prop_fam_of_interest
```


